\Large \bf Cyclone: Warp Speed Replication for Key Value Stores

### 2017-04-21 paper/paper.tex
Updates to paper.

\subsection{Core Allocation}
A key-value store integrated with Cyclone runs two types of threads. One type of
thread does key-value store work. This includes servicing lookup and update
requests and background threads that flush in-memory data structures to
secondary storage and do maintainence work on secondary storage data structures
such as compacting log-structured merge trees~\cite{lsmtree}. We use a dedicated
thread per-physical log to run the corresponding control plane including the
associated RAFT state machine. We use static assignment of threads to cores as
recommended for packet switching libraries such as DPDK for the best possible
performance. Therefore we use the term threads and cores interchangeably in this
paper. Client requests are always received by the control plane core for
the physical log the request is directed to. A response for the core is always
sent by the KV store core that finally handles the request.

An important resource for Cyclone is the NIC that must be shared among the
control plane cores that receive and replicate requests and the KV store cores
that send back responses. Most NICs provide independent transmit/receive queue
pairs and we aim for synchronization free access to these to aid horizontal
scalability with respect to the network resource. DPDK already provides fairly
scalable access to NIC queues, with the queue number being a parameter to send
and receive calls. All that remains for us is to map the queues to the control
plane and KV store cores, keeping in mind that there may be multiple NICs on the
system. We do this by iterating over all control plane cores followed by all KV
store cores. We assign two queue pairs to each control plane core, one to
receive requests from the client and the other transmit and receive replication
protocol related messages. We assign one queue pair to each KV store core to
send back responses to clients. We are careful to spread the different queue
pairs for a core across the available NICs as far as possible. This is important
due to the asymmetry between the different types of traffic. Queue pairs
carrying replication traffic will likely be more busy that those carrying client
request responses due to the amplification to multicast the same message to
different replicas.

The flow for an update operation in Cyclone is as follows. The network packet is
received by the control plane for the targeted physical log. The packet is then
immediately sent out for replication as described in the previous section. When
responses from a majority quorum is received, the request is passed to the
targeted KV store core for execution. The KV store core process the request and
sends back the response to the client. We note that replication is pipelined and
decoupled from execution and thus multiple updates requests are usually in
flight for replication at a time.

The flow for a read operation is simpler. It is still targeted to a physical log
but instead to replicating it the control plane core simply passes it on
straight away to the KV store core for execution and response.

The control plane cores and KV store cores communicate through a one-way
lock-free FIFO queue passing pointers to packets. We leverage the convenient
ring library in DPDK for this.

Finally an asynchronous source of work in the system is maintenance of the
two-level log. We depart from a 1:1 mapping between the first level NVM log and
the second level flashlog here. Each physical log consists of one NVM log
managed by the corresponding control plane core and multiple flashlogs each of
which is managed as an additional responsibility by a KV store core. We
therefore enforce a many to one mapping between key value store cores and
control plane cores - thereby partitioning KV store cores across control plane
cores. A request picks a control plane core and any one of the associated KV
store cores for execution. The request is logged in the NVM log of the control
plane core and garbage collected once it is drained to the flashlog of the
selected KV store core.

\subsection{Leveraging API Commutativity}
Update requests to a key-value store consist of a key and an associated operation -
such as ``put'' that set the value for the key or ``merge'' that calls a
user-defined operator to merge a new value for the key into the existing value.
A naive approach that maps a key to a random physical log and subsequently to a
random core on the system for execution runs into the determinism problem -
replicas could diverge with respect to the final value assigned to a key on
different replicas. We illustrate this problem with an example in
Figure~\ref{fig:race} where two updates to the same key race with each other on
different physical logs and execution cores to end up being applied in different
orders on different replicas, leading to a divergence of state.

The solution to this problem is observe that the key-value store is a concurrent
data structure that ultimately admits a serialization of operations to
it. Further, consecutive operations to different keys in the serial history can
be interchanged, because operations to different keys are
\emph{commutative}. The correct way to use different physical logs is therefore
to simply hash the key to select the physical log \emph{and} the core on which
the operation executes. This guarantees determinism as the serial history of
two different replicas can be transformed into the same serial history by
reordering operations to different keys.

Further relaxations to the mapping scheme are possible when operations to the
same key are also commutative. A good example is when key value stores are used
to maintain counters. Different increment operations to the same counter are
commutative if one is prepared to tolerate non-determinism in the sequence of
updates seen by a reader with the counter reaching the same eventual value. This
specific use-case, in fact, lead to the addition of the merge operation in
Rocksdb, a key value store that we use for evaluation. Merge operations
therefore could be dispatched to a random physical log with no harm. For this
paper however, we stick to the stricter requirement that operations to the same
key are not commutative. However, Cyclone's client side libraries leave it to
the user to specify the exact physical log and key-value thread to use to enable
more relaxed mappings when necessary.

We note that commutativity as assumed by Cyclone is an \emph{API level
  property}. There are no assumption as to how application cores synchronize to
access the shared key-value state or the underlying data structures for holding
key-value pairs.

@@

\subsection{Ganged Operations}
Key value stores (such as Rocksdb) usually provide a batched write operation
that atomically updates a set of key-value pairs. A batched write is often a key
primitive for building more sophisticated operations such as
transactions. Batched writes require special handling in Cyclone due to the
requirement that we route a write request by hashing the key to select both the
control plane core (physical log) as well as KV store core. A batched write has
multiple keys and therefore we need to synchronize the progress of the operation
across multiple physical logs (and associated replication quorums) as well as KV
store cores. In effect the serial history contains a grouped set of operation on
multiple keys spanning multiple cores as a single atomic operation. All serial
histories across replicas therefore continue to map to the same equivalent
serial history modulo reordering of operations to different keys.

A simple solution to this problem is to use a distributed transaction, using
two-phase commit to couple the otherwise independent replication quorums and KV
store cores. However this method has a number of problems. First, it introduces
an additional network round trip that is normally not necessary for a key value
store with a single shared memory image holding all the keys. Second, handling
failures with distributed transactions would have cause an explosion in
complexity in Cyclone, something we wished to avoid in the interests of
practical deployment to independent end users who need to understand and 
maintain Cyclone as part of their KV store infrastructure.

Our solution instead is to use a technique we termed ganged operations, based on
the observation that in the failure free case, executing a batched write across
multiple control plane cores and KV store cores does not look very much
different from a synchronous shared memory barrier. The challenge is to
effectively manage failures in replication without causing participating cores
to stop making progress. The first step is to reconfigure RAFTs leader election
protocol to ensure that the leaders for all the physical logs are co-located on
the same host. Clients always dispatch batched writes to a fixed control plane
core (called the injection core) which is then responsible for serially
forwarding the requests to the participating control plane cores (for
corresponding physical logs). The injection core uses DPDKs packet cloning
primitives to avoid dataplane work while forwarding copies of the same batched
write packet to multiple control plane cores. The injection core also adds a
``nonce'' - a unique timestamp to the packet. In addition the injection core
adds a unique view number to the packet corresponding to the term numbers of all
the participating RAFT instances (read from shared memory of the co-located leaders).
We describe the remaining flow for ganged operations in terms of the pseudocode
in Figure~\ref{fig:ganged_ops}. We defer discussion of how we generate the nonce
to later in this section. We also assume that a unique barrier is allocated in
shared memory for each ganged operation. We discuss later how this is done
without using dynamic allocation.

In the absence of failures Figure~\ref{fig:ganged_ops} is straightforward. One
replication is complete on all participating physical logs, the corresponding KV
store cores execute {\tt event\_replicated\_ganged\_op}. The use a bitmask for
the barrier setting the appropriate bit for themselves. Once all KV store cores
have arrived at the barrier the operation is executed. A distinguished leader
core (we use use the minumum numbered participating KV store core) is
responsible for sending back the response to the client. Two types of failure
need to be considered. First, one of the non-leader cores could fail. The
remaning cores would never progress past the barrier. To solve this the leader
core montors the remaning cores for their current view - which is the RAFT term
of the last executed log entry. If the view has moved past the view in the
packet, replication failed on that physical log. The leader then sets the bit
for the core that will no longer participate in the barrier to aid progress (a
technique we borrowed from lock-free data structures~\cite{lock-free}). It also
sets a failed flag to signal failure to the other participating cores. The
second failure case is that the leader core might not receive the operation if
it fails during replication from corresponding physical log. This case is
detected in a similar manner by the remaining cores by monitoring the published
view from the leader core in shared memory.

@@

The assumption in Figure~\ref{fig:ganged_ops} is that each ganged operation is
mapped to a unique barrier. We achieve this by using a fixed piece memory owned
by the leader core to hold the barrier and write the nonce to it in order to
indicate that the barrer is active. Non-leader cores watch for the nonce to know
when to execute the ganged operation barrier in Figure~\ref{fig:ganged_ops},
while also monitoring the leader's published view to detect the case where the
ganged operation fails to replicate on the leader's physical log.

Finally, we describe how we generate the nonce. The nonce is generated on the
injector core by concatenating the ethernet MAC ID of the first NIC on the
system with a 64 bit value that is the number of CPU timestamp counter cycles
since epoch time (read from the real time clock at startup plus the number of
cycles from the CPU {\tt rdtsc} instruction). The nonce can only be repeated if
the same machine manages to fail and come back up in less time than the real
time clock drift (controlled with NTP), a possiblity that we discount.

Ganged operations possibly consititute the most complex part of Cyclone but the
code weighs in at well under a couple of hundred lines of code. We believe that
this complexity is worth paying to avoid distributed transactions.

>>>>

\subsection{Core Allocation}
A key-value store integrated with Cyclone runs two types of threads. The first
type of thread does key-value store work. This includes servicing lookup and
update requests and background threads that flush in-memory data structures to
secondary storage and do maintainence work on secondary storage data structures
such as compacting log-structured merge trees~\cite{lsmtree}. The second type of
thread is a dedicated thread per-physical log to run the corresponding control
plane including the associated RAFT state machine. We use static assignment of
threads to cores as recommended for packet switching libraries such as DPDK for
the best possible performance. Therefore we use the terms thread and core
interchangeably in this paper. Client requests are always received by the
control plane core for the physical log the request is directed to. A response
for the core is always sent by the KV store core that finally handles the
request.

An important resource for Cyclone is the NIC that must be shared among the
control plane cores that receive and replicate requests and the KV store cores
that send back responses. Most NICs provide independent transmit/receive queue
pairs and we aim for synchronization free access to these to aid horizontal
scalability with respect to the network resource. DPDK already provides fairly
scalable access to NIC queues, with the queue number being a parameter to send
and receive calls. All that remains for us is to map the queues to the control
plane and KV store cores, keeping in mind that there may be multiple NICs on the
system. We do this by iterating over all control plane cores followed by all KV
store cores. We assign two queue pairs to each control plane core, one to
receive requests from the client and the other to transmit and receive
replication protocol related messages. We assign one queue pair to each KV store
core to send back responses to clients. We are careful to spread the different
queue pairs for a core across the available NICs as far as possible. This is
important due to the asymmetry between the different types of traffic. Queue
pairs carrying replication traffic will likely be more busy than those carrying
client request responses due to the amplification in message count to multicast
the same message to different replicas.

The flow for an update operation in Cyclone is as follows. The network packet is
received by the control plane for the targeted physical log. The packet is then
immediately sent out for replication as described in the previous section. When
responses from a majority quorum is received, the request is passed to the
targeted KV store core for execution. The KV store core process the request and
sends back the response to the client. We note that replication is pipelined and
decoupled from execution and thus multiple updates are usually in
flight on the network for replication at the same time.

The flow for a read operation is simpler. It is still targeted to a physical log
but instead of replicating it, the control plane core simply passes it on
straight away to the KV store core for execution and response.

The control plane cores communicate with the KV store cores through a one-way
lock-free FIFO queue passing pointers to packets. We leverage the convenient
ring library in DPDK for this.

Finally, an asynchronous source of work in the system is maintenance of the
two-level log. We depart from a 1:1 mapping between the first level NVM log and
the second level flashlog here. Each physical log consists of one NVM log
managed by the corresponding control plane core and multiple flashlogs each of
which is managed as an additional responsibility by a KV store core. We
therefore enforce a many to one mapping between key value store cores and a
front end control plane core - thereby partitioning KV store cores across
control plane cores. A request picks a control plane core and any one of the
associated KV store cores for execution. The request is logged in the NVM log of
the control plane core and garbage collected once it is drained to the flashlog
of the selected KV store core.

\subsection{Leveraging API Commutativity}
Update requests to a key-value store include a key and an associated operation -
such as ``put'' that sets the value for the key or ``merge'' that calls a
user-defined operator to merge a new value for the key into the existing value.
A naive approach that maps a key to a random physical log and subsequently to a
random core on the system for execution runs into the determinism problem -
replicas could diverge with respect to the final value assigned to a key on
different replicas. We illustrate this problem with an example in
Figure~\ref{fig:race} where two updates to the same key race with each other on
different physical logs and execution cores to end up being applied in different
orders on different replicas, leading to a divergence of state.

The solution to this problem is observe that the key-value store is a concurrent
data structure that ultimately admits a serialization of operations to
it. Further, consecutive operations to different keys in the serial history can
be interchanged, because operations to different keys are
\emph{commutative}. The correct way to use different physical logs is therefore
to simply hash the key to select the physical log \emph{and} the core on which
the operation executes. This guarantees determinism as the serial history of
two different replicas can be transformed into the same serial history by
reordering operations to different keys. Operations to the same key are already
in the same order in the serial histories by virtue of being replicated on the
same physical log and being executed on the same KV store core.

Further relaxations to the mapping scheme are possible when operations to the
same key are also commutative. A good example is when key value stores are used
to maintain counters. Different increment operations to the same counter are
commutative if one is prepared to tolerate non-determinism in the sequence of
updates seen by a reader with the counter reaching the same eventual value. This
specific use-case, in fact, lead to the addition of the merge operation in
Rocksdb, a key value store that we use for evaluation. Merge operations
therefore could be dispatched to a random physical log with no harm. For this
paper however, we stick to the stricter case that operations to the same
key are not commutative. However, Cyclone's client side libraries leave it to
the user to specify the exact physical log and key-value store core to use in
order to enable more relaxed mappings when necessary.

We note that commutativity as assumed by Cyclone is an \emph{API level
  property}. There are no assumption as to how application cores synchronize to
access the shared key-value state or the underlying data structures for holding
key-value pairs. We emphasize that this means replicas ultimately have the same
content but not necessarily the same memory state.

@@

\subsection{Ganged Operations}
Key value stores (such as Rocksdb) usually provide a batched write operation
that atomically updates a set of key-value pairs. A batched write is often a key
primitive for building more sophisticated operations such as
transactions. Batched writes require special handling in Cyclone due to the
requirement that we route a write request by hashing the key to select both the
control plane core (physical log) as well as KV store core. A batched write has
multiple keys and therefore we need to synchronize the progress of the operation
across multiple physical logs (and associated replication quorums) as well as KV
store cores. In effect, the serial history contains a grouped set of operation
on multiple keys as a single atomic operation. All serial histories across
replicas therefore continue to map to the same equivalent serial history modulo
reordering of operations to different keys.

One way to do this is to use a distributed transaction, using
two-phase commit to couple the otherwise independent replication quorums and KV
store cores. However this method has a number of problems. First, it introduces
an additional network round trip that is normally not necessary for a key value
store with a single shared memory image holding all the keys. Second, handling
failures with distributed transactions would have caused an explosion in
complexity in Cyclone, something we wished to avoid in the interests of
practical deployment to independent end users who need to understand and 
maintain Cyclone as part of their embedded KV store infrastructure.

Our solution instead is to use a technique we termed ganged operations. It is
based on the observation that in the failure free case, executing a batched
write across multiple control plane cores and KV store cores does not look very
much different from a synchronous shared memory barrier. The challenge is to
effectively manage failures in replication without causing participating cores
to stop making progress. The first step is to reconfigure RAFTs leader election
protocol to ensure that the leaders for all the physical logs are co-located on
the same host. Clients always dispatch batched writes to a fixed control plane
core (called the injection core) which is then responsible for forwarding the
request to the participating control plane cores (for corresponding physical
logs). The injection core uses DPDKs packet cloning primitives to avoid
dataplane work while forwarding copies of the same batched write packet to
multiple control plane cores. The injection core also adds a ``nonce'' - a
unique timestamp to the packet. In addition, the injection core adds a unique
view number to the packet containing the term numbers of all the participating
RAFT instances (read from shared memory of the co-located leaders).  We describe
the remaining flow for ganged operations in terms of the pseudocode in
Figure~\ref{fig:ganged_ops}. We defer discussion of how we generate the nonce to
later in this section. We also assume that a unique barrier is allocated in
shared memory for each ganged operation. We discuss later how this is done
without using dynamic allocation.

In the absence of failures Figure~\ref{fig:ganged_ops} is straightforward. Once
replication is complete on all participating physical logs, the corresponding KV
store cores execute {\tt event\_replicated\_ganged\_op}. They use a bitmask for
the barrier setting the appropriate bit for themselves. Once all KV store cores
have arrived at the barrier the operation is executed. A distinguished leader
core (we use the minumum numbered participating KV store core) is
responsible for sending back the response to the client. Two types of failure
need to be considered. First, replication on the physical log corresponding to
one of the non-leader cores could fail. The
remaning cores would never progress past the barrier. To solve this the leader
core montors the remaning cores for their current view - which is the RAFT term
of the last executed log entry. If the view has moved past the view in the
packet, replication failed on that physical log. The leader then sets the bit
for the core that will no longer participate in the barrier to aid progress (a
technique we borrowed from lock-free data structures~\cite{lock-free}). It also
sets a failed flag to signal failure to the other participating cores. The
second failure case is that the leader core might not receive the operation if
it fails during replication from corresponding physical log. This case is
detected in a similar manner by the remaining cores by monitoring the published
view from the leader core in shared memory. We add that on a leadership change
due to timeout we enqueue and execute a nop on the physical log and all
associated KV store cores to ensure previous entries are committed - a corner
case in the RAFT protocol.

@@

The assumption in Figure~\ref{fig:ganged_ops} is that each ganged operation is
mapped to a unique barrier. We achieve this by using a fixed piece of memory
owned by the leader core to hold the barrier and write the nonce to it in order
to indicate that the barrer is active for the corresponding batched write
operation. Non-leader cores watch for the nonce to know when to execute the
ganged operation barrier in Figure~\ref{fig:ganged_ops}, while also monitoring
the leader's published view to detect the case where the ganged operation fails
to replicate on the leader's physical log.

Finally, we describe how we generate the nonce. The nonce is generated on the
injector core by concatenating the ethernet MAC ID of the first NIC on the
system with a 64 bit value that is the number of CPU timestamp counter cycles
since epoch time (read from the real time clock at startup plus the number of
cycles from the CPU {\tt rdtsc} instruction). The nonce can only be repeated if
the same machine manages to fail and come back up in less time than the real
time clock drift (controlled with NTP), a possibility that we discount.

Ganged operations possibly consititute the most complex part of Cyclone but the
code weighs in at well under a couple of hundred lines. We believe that
this additional complexity is still small compared to distributed transactions.
